\documentclass{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{geometry}
\geometry{margin=1in}

% Theorem environments
\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}{Proposition}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{remark}{Remark}[section]

% Code listing style for TAGL
\lstdefinelanguage{TAGL}{
  keywords={},
  sensitive=false,
  comment=[l]{--},
  morecomment=[s]{-*}{*-},
  morestring=[b]",
}
\lstdefinestyle{tagl}{
  language=TAGL,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue},
  commentstyle=\color{gray}\itshape,
  stringstyle=\color{teal},
  backgroundcolor=\color{gray!8},
  frame=single,
  framesep=4pt,
  rulecolor=\color{gray!40},
  breaklines=true,
  captionpos=b,
}

% Code listing style for C++
\lstdefinestyle{cpp}{
  language=C++,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue},
  commentstyle=\color{gray}\itshape,
  stringstyle=\color{teal},
  backgroundcolor=\color{gray!8},
  frame=single,
  framesep=4pt,
  rulecolor=\color{gray!40},
  breaklines=true,
  captionpos=b,
}

\title{\textbf{The Mathematics of tagd}\\
\large A Formal Description of the Mathematical Structures\\
in the tagd Semantic-Relational System}
\author{Isaac N Colley}
\date{April 2026}

\begin{document}

\maketitle

\begin{abstract}
The tagd system was built from engineering instincts about how knowledge should
be structured. This paper makes the mathematics of those instincts explicit.
Every claim here is grounded in the tagd source code and the TAGL language
specification. The core thesis is simple: \emph{information is structural}. A
tag's meaning is not intrinsic --- it is determined entirely by its position in
the tagspace, by its relations to other tags. This is not a metaphor. It is a
precise mathematical claim, and the sections below give it precise mathematical
form, identifying in the source code the structures of set theory, partial
orders, and Alexandrov topology.
\end{abstract}

\tableofcontents
\bigskip

\section{Primitives}

\subsection{Tags}

A \textbf{tag} is the atomic unit of the tagd system. All tokens in TAGL are
UTF-8 labels \cite{readme}. A tag carries two distinct identities:

\begin{itemize}
  \item \textbf{Nominal identity}: a globally unique UTF-8 label.
  \item \textbf{Structural identity}: its position in the tagspace hierarchy,
        given by its rank, defined by its subordinate relation (aka identity
        relation) \cite{spec}.
\end{itemize}

\begin{definition}[Tag Set]
Let $T$ be the set of all tags in a tagspace. Each tag $t \in T$ is identified
by a string $\mathrm{id}(t) \in \Sigma^*$, where $\Sigma^*$ denotes the set of
all finite UTF-8 strings.
\end{definition}

In TAGL, defining a tag looks like:
\begin{lstlisting}[style=tagl, caption={Defining a tag via a subordinate relation}]
>> dog _is_a mammal;
\end{lstlisting}
Here \texttt{dog} is a tag. Its nominal identity is the string \texttt{"dog"}.
Its structural identity is defined by its subordinate relation to
\texttt{mammal}.

\subsection{The Root and the Axiom}

Every tagspace has a distinguished root tag: \texttt{\_entity}. The README
states:
\begin{quote}
\texttt{\_entity} is axiomatic and self-referencing
(i.e.\ \texttt{\_entity \_sub \_entity}). It is the only tag having this
relation. It is the root of a tagspace.
\end{quote}

\begin{definition}[Axiomatic Root]
The tag \texttt{\_entity} is the \textbf{axiomatic root} of every tagspace. It
is the unique fixed point of the subordinate relation: the one tag whose
super-object is itself.
\end{definition}

In TAGL notation:
\begin{lstlisting}[style=tagl, caption={The axiomatic root}]
_entity _sub _entity
\end{lstlisting}

In set-theoretic terms, \texttt{\_entity} is the \textbf{least element} of the
tagspace under the subordinate order --- the tag that is an ancestor of
everything, including itself.

\section{The Subordinate Relation and the Rooted Tree}

\subsection{Subordinate Relations}

A tag's structural identity is defined by exactly one subordinate (sub)
relation. The TAGL specification states \cite{spec}:
\begin{quote}
A subject MUST have exactly one subordinate relation to exist in the tagspace.
\end{quote}

The sub relators are defined in \texttt{hard-tags.h} \cite{hardtags}:
\begin{lstlisting}[style=cpp, caption={Sub relator definitions in hard-tags.h}]
#define HARD_TAG_SUB     "_sub"     // most abstract sub relator
#define HARD_TAG_IS_A    "_is_a"    // instance subordination
#define HARD_TAG_TYPE_OF "_type_of" // type subordination
\end{lstlisting}

A typical TAGL knowledge hierarchy:
\begin{lstlisting}[style=tagl, caption={Building a subordinate hierarchy in TAGL}]
>> physical_object -^ _entity;
>> living_thing _type_of physical_object;
>> animal _type_of living_thing;
>> mammal _type_of animal;
>> dog _is_a mammal;
\end{lstlisting}

\subsection{The Tagspace as a Rooted Tree}

The subordinate relation organizes all tags into a \textbf{rooted tree}.

\begin{definition}[Tagspace Tree]
A tagspace is a pair $(T, p)$ where:
\begin{itemize}
  \item $T$ is a set of tags containing the distinguished root
        \texttt{\_entity},
  \item $p : T \setminus \{\texttt{\_entity}\} \to T$ is the \emph{parent
        function}, assigning each non-root tag its super-object.
\end{itemize}
There are no cycles (except the axiomatic self-reference of
\texttt{\_entity}).
\end{definition}

The tree structure of the example above:
\begin{verbatim}
_entity
  physical_object
    living_thing
      animal
        mammal
          dog
\end{verbatim}

This is not merely a data structure choice. The tree encodes \emph{taxonomic
knowledge}: to be a \texttt{dog} is to be a \texttt{mammal}, which is to be an
\texttt{animal}, and so on. Structural position \emph{is} meaning.

\section{Rank: Encoding the Tree in Strings}

\subsection{The Rank Function}

Each tag is assigned a \textbf{rank} --- a UTF-8 byte string encoding its
position in the tagspace tree. The \texttt{rank.h} source states:
\begin{quote}
A rank string is the ordinal value of each node in the tagspace tree. A code
point at a particular level is known as the rank value.
\end{quote}

\begin{definition}[Rank Function]
The rank function $r : T \to \Sigma^*$ assigns to each tag a UTF-8 string
such that the string's structure reflects the tag's position in the tree.
\end{definition}

Concrete rank assignments from the bootstrap dump \cite{bootstrap}:

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Tag} & \textbf{Rank (dotted)} & \textbf{Position} \\
\midrule
\texttt{\_entity}       & (empty)  & level 0, root \\
\texttt{\_sub}          & 1        & level 1 \\
\texttt{\_is\_a}        & 1.1      & level 2 \\
\texttt{\_type\_of}     & 1.2      & level 2 \\
\texttt{\_rel}          & 2        & level 1 \\
\texttt{\_has}          & 2.1      & level 2 \\
\texttt{\_number}       & 3        & level 1 \\
\texttt{\_integer}      & 3.1      & level 2 \\
\texttt{\_float}        & 3.2      & level 2 \\
\texttt{\_interrogator} & 4        & level 1 \\
\bottomrule
\end{tabular}
\end{center}

\subsection{The Prefix Partial Order}

The key property of ranks is stated explicitly in \texttt{rank.h}:
\begin{quote}
All children of a particular node in the rank hierarchy will be prefixed by the
parent node --- that is, rank substrings define subtrees in the hierarchy.
\end{quote}

\begin{definition}[Containership Relation]
Define the \textbf{containership relation} $\sqsubseteq$ on $T$ by:
\[
  a \sqsubseteq b
  \iff
  r(b) \text{ is a prefix of } r(a).
\]
That is, $a \sqsubseteq b$ means $b$ is an ancestor of $a$ (or $b = a$). Tag
$b$ \emph{contains} tag $a$ in the hierarchy.
\end{definition}

This relation is implemented directly in \texttt{rank.h}:
\begin{lstlisting}[style=cpp, caption={Containership via rank prefix in rank.h}]
// this rank contains the rank being compared,
// in other words, this rank is a prefix to the other
bool contains(const rank&) const;
\end{lstlisting}

And used in \texttt{tagd.cc} for structural operations on tag sets:
\begin{lstlisting}[style=cpp, caption={Containership used in merge\_containing\_tags (tagd.cc)}]
if (b->rank().contains(a->rank())) { // rank can contain itself (b == a)
    merged++;
} else if (a->rank().contains(b->rank())) {
    A.insert(*b);
}
\end{lstlisting}

\subsection{Verifying the Partial Order Axioms}

\begin{theorem}
The containership relation $\sqsubseteq$ is a partial order on $T$.
\end{theorem}

\begin{proof}
We verify the three required properties.

\textbf{Reflexivity.} \texttt{rank.h} notes that \texttt{contains()} holds when
$b = a$, since every string is a prefix of itself. Hence $a \sqsubseteq a$
for all $a \in T$.

\textbf{Antisymmetry.} Suppose $a \sqsubseteq b$ and $b \sqsubseteq a$. Then
$r(b)$ is a prefix of $r(a)$ and $r(a)$ is a prefix of $r(b)$. A string $u$
is a prefix of $v$ and $v$ is a prefix of $u$ if and only if $u = v$. Hence
$r(a) = r(b)$, which implies $a = b$ since ranks uniquely identify tags.

\textbf{Transitivity.} Suppose $a \sqsubseteq b$ and $b \sqsubseteq c$. Then
$r(b)$ is a prefix of $r(a)$ and $r(c)$ is a prefix of $r(b)$. Since the
prefix relation on strings is transitive, $r(c)$ is a prefix of $r(a)$, so
$a \sqsubseteq c$.
\end{proof}

Therefore $(T, \sqsubseteq)$ is a \textbf{poset} (partially ordered set).

\subsection{Lexicographic Total Order}

Ranks also carry a total \textbf{lexicographic order}, implemented in
\texttt{rank.h}:
\begin{lstlisting}[style=cpp, caption={Lexicographic ordering of ranks}]
bool operator<(const rank& rhs) const { return _data < rhs._data; }
\end{lstlisting}

This order --- a deliberate UTF-8 design choice noted in \texttt{rank.h}
(``it collates lexicographically'') --- enables efficient storage and retrieval
of the entire tagspace as a sorted structure. Tags within the same subtree sort
together. The tree \emph{is} the index.

\section{The Tagspace Topology}

\subsection{From Partial Order to Topology}

Every partial order canonically generates a topology. Given the poset
$(T, \sqsubseteq)$, we define:

\begin{definition}[Tagspace Open Sets]
A subset $U \subseteq T$ is \textbf{open} if it is upward-closed under
$\sqsubseteq$:
\[
  U \in \mathcal{T}
  \iff
  \forall\, a \in U,\; \forall\, b \in T :\;
  a \sqsubseteq b \Rightarrow b \in U.
\]
\end{definition}

In the tagspace tree, ``upward-closed'' means: if a tag is in the set, all
its ancestors are too. Each basic open set is the \textbf{subtree rooted at a
tag} together with all tags above it.

This construction is known as the \textbf{Alexandrov topology} --- the topology
generated by a partial order in which arbitrary intersections (not just finite
ones) of open sets remain open.

\subsection{Verifying the Topology Axioms}

\begin{theorem}
Let $M$ be the set of all tags in a tagspace and let $\mathcal{T}$ be the
collection of upward-closed subsets under $\sqsubseteq$. Then
$(M, \mathcal{T})$ is a topological space.
\end{theorem}

\begin{proof}
We verify the three axioms.

\textbf{(i) $\emptyset \in \mathcal{T}$ and $M \in \mathcal{T}$.}
The empty set is vacuously upward-closed. The full set $M$ is upward-closed
because there is no $b \in T$ outside $M$. Both belong to $\mathcal{T}$.

\textbf{(ii) Finite intersections.}
Let $U, V \in \mathcal{T}$. Suppose $a \in U \cap V$ and $a \sqsubseteq b$.
Then $b \in U$ (since $U$ is upward-closed) and $b \in V$ (since $V$ is
upward-closed), so $b \in U \cap V$. Hence $U \cap V \in \mathcal{T}$.

\textbf{(iii) Arbitrary unions.}
Let $\{U_\alpha\}$ be any collection of sets in $\mathcal{T}$. Suppose
$a \in \bigcup_\alpha U_\alpha$ and $a \sqsubseteq b$. Then $a \in U_\alpha$
for some $\alpha$. Since $U_\alpha$ is upward-closed, $b \in U_\alpha
\subseteq \bigcup_\alpha U_\alpha$. Hence $\bigcup_\alpha U_\alpha \in
\mathcal{T}$.
\end{proof}

Therefore $(M, \mathcal{T})$ is a \textbf{topological space}. The
\texttt{merge\_containing\_tags} function in \texttt{tagd.cc} computes
exactly the intersection of two subtrees, exhibiting axiom (ii) in code:
\begin{lstlisting}[style=cpp, caption={Intersection of subtrees in tagd.cc}]
if (b->rank().contains(a->rank())) {
    merged++;
} else if (a->rank().contains(b->rank())) {
    A.insert(*b);
}
\end{lstlisting}

\subsection{Specialization Order}

In the Alexandrov topology, every topological space carries a natural
\textbf{specialization preorder}: $a$ specializes $b$ (written
$a \rightsquigarrow b$) when $a$ is in the topological closure of $\{b\}$.

\begin{proposition}
In the tagspace topology, the specialization order coincides with the
subordinate order: $a \rightsquigarrow b$ if and only if $a$ is a descendant
of $b$ in the tagspace tree.
\end{proposition}

In plain language: \texttt{dog} specializes \texttt{mammal}, which specializes
\texttt{animal}. More specific tags (deeper in the tree) specialize more general
ones. Taxonomic inheritance is not a rule to be enforced --- it is a geometric
consequence of the shape of the space.

\subsection{Continuity as Meaning-Preservation}

\begin{definition}[Tagspace Map]
A map $f : T_1 \to T_2$ between two tagspaces is \textbf{topologically
continuous} if it preserves open sets: $f^{-1}(U) \in \mathcal{T}_1$ for
every $U \in \mathcal{T}_2$.
\end{definition}

\begin{proposition}
A map $f : T_1 \to T_2$ is continuous (with respect to the Alexandrov
topologies) if and only if it is order-preserving: whenever
$a \sqsubseteq b$ in $T_1$, we have $f(a) \sqsubseteq f(b)$ in $T_2$.
\end{proposition}

This gives a precise mathematical definition of a
\textbf{meaning-preserving transformation} between tagspaces. If
\texttt{dog \_is\_a mammal} in $T_1$, a continuous map must send \texttt{dog}
somewhere below the image of \texttt{mammal} in $T_2$.

\section{The Hard Tag Base: A Universal Subspace}

\subsection{Hard Tags}

Every tagspace --- no matter how different its user-defined tags --- contains
exactly the same set of hard tags, defined in \texttt{hard-tags.h}
\cite{hardtags}:

\begin{lstlisting}[style=cpp, caption={Core hard tag definitions in hard-tags.h}]
#define HARD_TAG_ENTITY       "_entity"       // root
#define HARD_TAG_SUB          "_sub"          // subordination
#define HARD_TAG_IS_A         "_is_a"         // instance relation
#define HARD_TAG_TYPE_OF      "_type_of"      // type relation
#define HARD_TAG_RELATOR      "_rel"          // predicate root
#define HARD_TAG_HAS          "_has"          // possession
#define HARD_TAG_CAN          "_can"          // capability
#define HARD_TAG_NUMBER       "_number"       // numeric type root
#define HARD_TAG_INTEGER      "_integer"      // integer type
#define HARD_TAG_FLOAT        "_float"        // float type
#define HARD_TAG_INTERROGATOR "_interrogator" // query root
\end{lstlisting}

\subsection{The Hard Tag Subspace}

\begin{definition}[Hard Tag Subspace]
Let $H \subset T$ denote the set of hard tags in any tagspace $T$. The
induced subspace topology on $H$ is identical in every tagspace. $H$ is
therefore the \textbf{universal invariant subspace} --- a fixed topological
base that every tagspace shares by construction.
\end{definition}

\begin{remark}
In categorical terms, the hard tag subspace is the \textbf{initial object} in
the category of tagspaces: every tagspace is an extension of $H$, and every
valid map between tagspaces must fix $H$.
\end{remark}

\subsection{Interoperability via the Common Base}

Because every tagspace contains $H$ as a common topological foundation, any
two tagspaces can communicate. The hard tags are the shared language --- the
invariant ground that makes the \texttt{httagd} web protocol possible. Two
tagspaces may diverge arbitrarily in their user-defined tags, but they always
agree on the structure of $H$.

\begin{lstlisting}[style=tagl, caption={Hard tags are invariant; user tags vary}]
-- invariant in every tagspace:
_entity _sub _entity

-- user-defined, varies by tagspace:
>> dog _is_a mammal;
\end{lstlisting}

\section{Predicates: Fibered Structure}

\subsection{Relators Are Tags}

Crucially, relators are themselves tags --- subordinate to \texttt{\_rel} in
the tagspace tree. This means predicates are not outside the system; they are
\emph{inside} it. The tagspace is self-describing.

From the bootstrap dump \cite{bootstrap}:
\begin{verbatim}
_rel     rank: 2        (root of all relators)
_has     rank: 2.1
_can     rank: 2.2
_refers  rank: 2.3
\end{verbatim}

The language used to attach meaning to tags is itself expressed in the same
structural language.

\subsection{Predicates as Horizontal Relations}

Predicates are relations attached to a tag that do not alter its position in
the tree. From \texttt{TAGL-spec.md} \cite{spec}:
\begin{quote}
A predicate consists of a relator followed by an object\_list. A relator MUST
be a tag subordinate to \texttt{\_rel}.
\end{quote}

\begin{lstlisting}[style=tagl, caption={Vertical identity and horizontal predicates}]
>> dog _is_a mammal    -- vertical: places dog in the tree
    _has legs = 4, tail
    _can bark;         -- horizontal: predicate data at dog
\end{lstlisting}

\subsection{The Two-Layer Structure}

The tagspace carries two distinct mathematical layers:

\begin{definition}[Two-Layer Tagspace Structure]
\begin{enumerate}
  \item \textbf{Base space}: the topological space $(M, \mathcal{T})$ defined
        by subordinate relations and ranks --- the vertical dimension.
  \item \textbf{Fiber}: at each tag $t \in M$, a set of predicates
        $P(t) \subseteq \mathrm{Rel} \times T \times \mathrm{Mod}$ --- the
        horizontal dimension, where $\mathrm{Rel}$ is the set of relator tags
        and $\mathrm{Mod}$ is the set of modifier values.
\end{enumerate}
\end{definition}

This two-layer structure --- a base space with additional data attached at each
point --- is analogous to a \textbf{fiber bundle} in differential geometry. The
topology gives the skeleton; the predicates give the flesh.

The \texttt{abstract\_tag} class in \texttt{tagd.h} encodes exactly this
decomposition:
\begin{lstlisting}[style=cpp, caption={Two-layer structure in abstract\_tag (tagd.h)}]
class abstract_tag {
    tagd::rank    _rank;       // position in the base space
    predicate_set _relations;  // fiber: additional data at this point
};
\end{lstlisting}

\section{Structural Identity}

\begin{center}
    \textit{``To be, is to be related.''} \\[0.5ex]
    \small --- Maxim of TAGL
\end{center}

\subsection{Identity Is Position}

The deepest mathematical claim of the tagd system is that a tag's identity is
entirely structural. The TAGL specification defines:
\begin{quote}
\textbf{Structural identity}: its position in the tagspace hierarchy, given by
its rank, defined by its subordinate relation (aka identity relation).
\end{quote}

A tag \texttt{dog} is not \texttt{dog} because of the string \texttt{"dog"}.
It is \texttt{dog} because of where it sits --- under \texttt{mammal}, under
\texttt{animal}, under \texttt{living\_thing}, under \texttt{\_entity} --- and
because of what predicates are attached to it there. Change the position (the
subordinate relation), and you have a different tag, regardless of the label.

This is \textbf{mathematical structuralism} applied to knowledge
representation: identity is relational, not intrinsic.

\subsection{The Canonical Form}

A tagspace can be fully serialized as TAGL text via \texttt{dump()}. From
the TAGL Thesis document:
\begin{quote}
The canonical form of a tagspace is its dump --- the UTF-8 TAGL code that
represents the tagspace (and can be loaded).
\end{quote}

The dump is not a record of a structure. It \emph{is} the structure, expressed
in TAGL. The tagspace is its own specification.

\section{Deeper Structures}

The foundations established in Sections 1--8 are sufficient to describe what
the tagd system \emph{is}. This section describes what it \emph{implies} ---
deeper mathematical structures that emerge from those foundations and illuminate
how the system behaves.

\subsection{The Adjunction Between Order and Topology}

The relationship between the tagspace as a partial order and the tagspace as a
topological space is not incidental. It is a formal \textbf{adjunction} --- one
of the most fundamental relationships in category theory.

Two functors mediate between the two views:

\begin{itemize}
  \item $A : \mathbf{PreOrd} \to \mathbf{Top}$ (the \emph{Alexandrov functor})
        takes the rank prefix partial order on tags and produces the Alexandrov
        topological space described in Section~4.
  \item $S : \mathbf{Top} \to \mathbf{PreOrd}$ (the \emph{specialization
        functor}) takes any topological space and recovers a preorder:
        $a \leq b$ iff every open set containing $a$ also contains $b$.
\end{itemize}

These functors form an adjunction: $A$ is left adjoint to $S$, written
$A \dashv S$. Applying both in sequence recovers the original structure:
$S(A(P)) \cong P$.

In tagd, this adjunction is visible in the two-step construction of a fully
hydrated tag. A tag begins as a semantic object --- identity, subordinate
relation, position --- living in the preorder. When its rank is assigned, it is
lifted into the Alexandrov topological space. The rank upgrade is the
\textbf{unit of the adjunction}: the reflection morphism that moves a tag from
the order-theoretic world into the topological world while preserving its
identity on the underlying set.

\begin{lstlisting}[style=cpp, caption={The reflection morphism: lifting a tag from PreOrd into the Alexandrov space}]
// Semantic tag: lives in PreOrd
tagd::abstract_tag semantic(id, sub_relator, super_object, pos);

// Reflection map: lifts into the Alexandrov space
tagd::abstract_tag hydrated = rank.empty()
    ? std::move(semantic)
    : tagd::abstract_tag(semantic, rank);  // <- unit of the adjunction
\end{lstlisting}

The reflection morphism is:
\begin{itemize}
  \item \textbf{Bijective} on the underlying set of tags (same identity, same
        element)
  \item \textbf{Continuous} with respect to the Alexandrov topology
  \item \textbf{Order-reflecting} (preserves and reflects the subordinate
        relation via rank prefixes)
\end{itemize}

This is a \textbf{bimorphism} in the category of preorders and topological
spaces.

\subsection{Topological Properties of the Tagspace}

Because the tagspace carries an Alexandrov topology, it inherits a family of
topological properties that have direct operational consequences.

\textbf{First countable.} Every tag $x$ has a smallest open neighbourhood: the
set of all tags whose rank has $r(x)$ as a prefix --- all descendants of $x$
plus $x$ itself. This smallest neighbourhood is the local base at $x$. It is
why rank-based traversal (\texttt{get\_children}, \texttt{related}) is
efficient: the topology guarantees a finite, well-defined local structure at
every point.

\textbf{Locally compact.} For any tag $x$, the smallest open neighbourhood
$U(x)$ is compact in the Alexandrov topology. Any open cover of $U(x)$ must
contain $U(x)$ itself, giving a trivial finite subcover. This is why queries
over subtrees are well-behaved: every neighbourhood is compact by construction.

\textbf{Locally path connected.} Between any two comparable tags
$x \sqsubseteq y$, there is a path up the subordinate chain via successive
super-object steps. The smallest neighbourhood of any tag is connected (it is
an upper set), so the space is locally path-connected in the order-theoretic
sense.

\textbf{Subspaces and products are Alexandrov-discrete.} Any subset of tags ---
for example, the result of a \texttt{query()} or all descendants of a given tag
--- inherits the Alexandrov structure. Query results compose correctly because
they are subspaces of the same topological space, not alien data structures.

\subsection{Modal Operators: Interior and Closure}
\label{sec:modal}

In any topological space, the \textbf{interior} and \textbf{closure} operators
are well-defined. In the tagspace, they carry semantic meaning.

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Topological operator} & \textbf{Modal logic} & \textbf{tagd meaning} \\
\midrule
Interior $\mathrm{int}(U)$ & Necessarily ($\square$) & Properties true in all superiors of a tag \\
Closure $\mathrm{cl}(U)$ & Possibly ($\lozenge$) & Properties reachable via any subordinate \\
\bottomrule
\end{tabular}
\end{center}

\bigskip

The tagspace is therefore an \textbf{interior algebra} --- a Boolean algebra
equipped with an interior operator satisfying the Kuratowski axioms. The
\texttt{related()}, \texttt{has\_relator()}, and \texttt{get\_children()}
families of functions are the operational realization of these modal operators.

Furthermore, because the Alexandrov topology arises from a partial order, the
tagspace constitutes a \textbf{Kripke frame}: a set of worlds (tags) equipped
with an accessibility relation (the subordinate order). A predicate $P$ is
\emph{necessarily true} at tag $x$ (written $\square P(x)$) if $P$ holds at
every tag above $x$. It is \emph{possibly true} ($\lozenge P(x)$) if $P$
holds at some tag below $x$.

\begin{lstlisting}[style=tagl, caption={Modal reading of TAGL predicates}]
-- "dog necessarily _has legs" if all ancestors of dog also _has legs
-- "animal possibly _can bark" if some descendant of animal _can bark
\end{lstlisting}

This gives TAGL queries a natural modal interpretation without any additional
machinery. The topology provides the modal semantics for free.

\subsection{Constructive Existence: Tags as Witnesses}
\label{sec:constructive}

The trail from Kripke semantics through intuitionistic logic leads to a
foundational observation about how tagd represents knowledge.

In \textbf{classical mathematics}, an existence proof may assert ``there exists
an object with property $P$'' without producing one. In \textbf{constructive
mathematics}, existence requires a witness --- an explicit object that
demonstrates the property.

The tagd system is constructive by design. A tag does not \emph{exist} in the
tagspace until it is explicitly constructed with a subordinate relation. The
TAGL rule --- \emph{a tag MUST have exactly one subordinate relation} --- is a
constructive existence condition. No witness, no tag.

After the const-correctness work (Phase 9B), this constructive character is
enforced in the type system: tags are built once, immutably, in a single
construction step. There is no ``assert existence, hydrate later'' pattern. The
construction \emph{is} the proof of existence.

\begin{lstlisting}[style=cpp, caption={One-shot construction as constructive existence}]
// Constructive: the tag is its own existence proof.
// No tag exists until this line executes.
tagd::abstract_tag t(id, sub_relator, super_object, pos);
\end{lstlisting}

This aligns the implementation with \textbf{intuitionistic logic}, where a
proof of $\exists x.\, P(x)$ is a pair $(t, p)$ --- a term $t$ and a proof $p$
that $P(t)$ holds. In tagd: the term is the tag, and the proof is its rank and
subordinate relation.

\subsection{Martin-L\"{o}f Type Theory and Dependent Structure}
\label{sec:mltt}

The constructive character of tagd points toward \textbf{Martin-L\"{o}f Type
Theory} (MLTT) --- the most rigorous modern foundation for constructive
mathematics, and the theoretical basis for proof assistants such as Agda, Lean,
and Coq.

In MLTT, propositions are types and proofs are programs (the Curry-Howard
correspondence). Every existence claim must be witnessed by an explicit term
inhabiting the corresponding type.

The tagspace can be read as a fragment of MLTT:

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{MLTT concept} & \textbf{tagd realization} \\
\midrule
Type & Tag position in the tagspace (defined by rank) \\
Term inhabiting a type & A concrete tag at that position \\
Dependent type & Rank-dependent tag structure \\
Sigma type $\Sigma_{x:A}\, B(x)$ & A tag together with its predicate fiber \\
Identity type & Immutable structural identity: same rank $=$ same type \\
Universe & The tagspace as a whole \\
Constructive proof & One-shot tag construction; \texttt{dump()} as canonical form \\
\bottomrule
\end{tabular}
\end{center}

\bigskip

The rank function $r : T \to \Sigma^*$ is a \textbf{dependent type family}:
the structural type of a tag depends on its rank value. Two tags at different
ranks are, in the type-theoretic sense, of different types --- even if their
nominal labels are similar.

\begin{remark}
Making identity immutable in the C++ type system (Phase~9B) is not merely a
software quality improvement. It is a \textbf{formalization step}: the C++
types now enforce what MLTT would express as an identity type. Once a tag's
rank cannot be mutated, the fixed-point property of \texttt{\_entity} is not
just documented --- it is an axiom enforced by the compiler.
\end{remark}

\subsection{Computation as Topological Traversal}

The introduction of functions as tags in TAGL (Section~7 of the TAGL
specification) completes the arc from the tagspace as a knowledge structure
to the tagspace as a computing environment. The mathematical consequences
are significant.

\textbf{Functions are tags.} Defining a function is a \texttt{>>} PUT
operation --- a constructive existence claim. The function does not exist
until it is placed in the tagspace with a subordinate relation. It is subject
to the same identity rules, the same rank structure, and the same topological
properties as any other tag. It can be queried, inspected, versioned, and
compared.

\begin{lstlisting}[style=tagl, caption={A function defined as a tag via PUT}]
>> greet($name) _is_a function
    -> "Hello, " ++ $name;
\end{lstlisting}

This is not syntactic convenience. It means the computation is \emph{inside}
the topological space, not outside it.

\textbf{A function call is a topological operation.} A call navigates from
the function tag (a point in $T$) through its predicate fiber (the body) to
a result. Combined with the Curry-Howard correspondence from
Section~\ref{sec:mltt}: \emph{tags are types, functions are proofs, and
evaluation is topological traversal}.

\textbf{Single-assignment variables are proof terms.} Variables (\texttt{\$name})
follow single-assignment semantics --- immutable after binding. This is
directly consistent with the constructive / MLTT foundation of
Section~\ref{sec:constructive}: a variable binding is a term inhabiting a
type. Once constructed, it cannot be mutated, for the same reason a tag's
rank cannot be mutated after construction.

\textbf{Guards are modal assertions.} The \texttt{\_when} guard syntax
exposes the Kripke frame structure of the tagspace directly in the language.
A guard using $\square$ asserts necessity:

\begin{lstlisting}[style=tagl, caption={Modal guard using the necessity operator}]
>> area($shape) _when [](has radius) ->
    3.14159 * $shape.radius ** 2;
\end{lstlisting}

This asserts that the function applies only when \texttt{\_has radius} is
\emph{necessarily} true --- true at this tag and all its ancestors in the
Alexandrov topology. The modal operators $\square$ and $\lozenge$ are not
new machinery; they are the interior and closure operators of
Section~\ref{sec:modal}, exposed as first-class language constructs.

\textbf{Async processes communicate via continuous maps.} The Erlang-style
process model in TAGL sends TAGL text through buffers between isolated
processes. Topologically, a channel between two tagspace-aware processes is
a \emph{continuous map} --- an order-preserving transformation that preserves
the subordinate structure. The hard tag base $H$ is the common topological
ground that makes inter-process TAGL intelligible: both ends of a channel
share the universal invariant subspace, guaranteeing that the structural
meaning of hard-tag constructs is preserved across the channel boundary.

The three-part alignment is now complete:

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Layer} & \textbf{Structure} & \textbf{Mathematical object} \\
\midrule
Knowledge    & Tags and subordinate relations          & Alexandrov topological space \\
Meaning      & Predicates at each tag                 & Fiber over the base space \\
Computation  & Functions as tags, evaluation as traversal & Topological map + Kripke semantics \\
\bottomrule
\end{tabular}
\end{center}

\section{Summary}

Table~\ref{tab:summary} collects all mathematical structures identified in the
tagd source, including those from the deeper structures of Section~9.

\begin{table}[h]
\centering
\small
\begin{tabular}{lll}
\toprule
\textbf{Mathematical Structure} & \textbf{tagd Realization} & \textbf{Source} \\
\midrule
Set & All tags $T$ in a tagspace & \texttt{tagd.h: tag\_set} \\
Rooted tree & Subordinate relations from \texttt{\_entity} & \texttt{tagd-README.md} \\
Partial order (poset) & Rank prefix containership $\sqsubseteq$ & \texttt{rank.h: contains()} \\
Lexicographic total order & Rank byte string comparison & \texttt{rank.h: operator<} \\
Alexandrov topology & Upward-closed subtrees as open sets & \texttt{rank.h}, \texttt{tagd.cc} \\
Universal base subspace & Hard tag set $H$ & \texttt{hard-tags.h} \\
Fiber / attached structure & Predicate sets $P(t)$ at each tag & \texttt{tagd.h: predicate\_set} \\
Continuous map & Order-preserving tagspace map & \texttt{tagd.cc: merge\_containing\_tags} \\
Canonical serialization & \texttt{dump()} in TAGL canonical form & \texttt{TAGL-spec.md} \\
Adjunction ($A \dashv S$) & Rank upgrade as reflection morphism & \texttt{tagd.cc: abstract\_tag} ctor \\
First countable & Smallest open nbhd via rank prefix & \texttt{rank.h: contains()} \\
Locally compact & Subtree queries always well-behaved & \texttt{tagd.cc: get\_children} \\
Interior algebra / modal ops & \texttt{related()}, \texttt{has\_relator()} as $\square$/$\lozenge$ & \texttt{tagd.h} \\
Kripke frame & Subordinate order as accessibility relation & \texttt{rank.h}, \texttt{tagd.cc} \\
Constructive existence & One-shot tag construction as witness & \texttt{tagd.h: abstract\_tag} ctor \\
Dependent type (MLTT) & Rank-dependent tag structure & \texttt{rank.h}, \texttt{tagd.h} \\
Functions as tags & PUT-defined executable tags in tagspace & \texttt{TAGL-spec.md} \S7 \\
Computation as traversal & Function call as topological map through fiber & \texttt{TAGL-spec.md} \S7 \\
Modal guards & \texttt{\_when} $\square$/$\lozenge$ as interior/closure in syntax & \texttt{TAGL-spec.md} \S7.4 \\
\bottomrule
\end{tabular}
\caption{Mathematical structures and their tagd source realizations.}
\label{tab:summary}
\end{table}

\section{Conclusion}

The tagd system is not merely a database or a knowledge graph. It is a
\textbf{topological space of structured meaning}. Tags are points. Subordinate
relations give the space its shape. Ranks encode that shape as a partial order.
The Alexandrov topology makes the shape mathematically precise. Predicates
attach additional structure at each point as a fiber over the base topology.
And the hard tag base ensures that every tagspace shares the same foundational
geometry.

TAGL is the language for acting on this space: defining points, querying
regions, transforming structure. Every TAGL statement is a topological
operation.

When functions enter the tagspace as tags, computation becomes part of the
same structure. A function call is a traversal from a point in $T$ through
its predicate fiber to a result. Guards express modal assertions over the
Kripke frame. Async processes communicate via continuous maps across the
shared hard tag base. Knowledge, meaning, and computation are no longer
separate concerns --- they are three layers of the same topological object.

The Curry-Howard correspondence completes the picture: tags are types,
functions are proofs, and evaluation is topological traversal.

The vision that \emph{information is structural} --- that meaning comes from
shape, not from labels --- is not a design philosophy. It is a theorem about
how the system works.

\begin{thebibliography}{9}

\bibitem{hardtags}
tagd source code, \texttt{tagd/include/tagd/hard-tags.h}.
Hard tag definitions and axioms.

\bibitem{rankh}
tagd source code, \texttt{tagd/include/tagd/rank.h}.
Rank structure, prefix order, and \texttt{contains()} implementation.

\bibitem{tagdh}
tagd source code, \texttt{tagd/include/tagd.h}.
\texttt{abstract\_tag}, \texttt{predicate\_set}, and tag set operations.

\bibitem{tagdcc}
tagd source code, \texttt{tagd/src/tagd.cc}.
\texttt{merge\_containing\_tags} and merge operations.

\bibitem{spec}
\textit{TAGL Specification}, \texttt{TAGL-spec.md}.
Language specification and semantic model.

\bibitem{thesis}
\textit{TAGL Design Thesis}, \texttt{TAGL-Thesis.md}.
Language design document including function model and computation semantics.

\bibitem{readme}
\textit{tagd README}, \texttt{tagd-README.md}.
Working examples and TAGL tutorial.

\bibitem{bootstrap}
\textit{Bootstrap Dump Grid}, \texttt{bootstrap-dump-grid.txt}.
Concrete tagspace rank assignments.

\end{thebibliography}

\end{document}
